Nail: A practical interface generator for data formats

### 2015-08-10 paper/thesis/code/zip_out.tex
thesis

\begin{Verbatim}[commandchars=\\\{\},codes={\catcode`\$=3\catcode`\^=7\catcode`\_=8}]
\PY{k}{struct} \PY{n}{zip}\PY{p}{\PYZob{}}
  \PY{n}{end\PYZus{}of\PYZus{}directory} \PY{n}{contents}\PY{p}{;}
\PY{p}{\PYZcb{}}\PY{p}{;}
\PY{k}{struct} \PY{n}{end\PYZus{}of\PYZus{}directory}\PY{p}{\PYZob{}}
  \PY{k+kt}{uint16\PYZus{}t} \PY{n}{disks}\PY{p}{;}
  \PY{k+kt}{uint16\PYZus{}t} \PY{n}{directory\PYZus{}disk}\PY{p}{;}\PY{c+cm}{/*...*/}
  \PY{k}{struct} \PY{p}{\PYZob{}}
    \PY{n}{dir\PYZus{}fileheader}\PY{o}{*} \PY{n}{elem}\PY{p}{;}
    \PY{k+kt}{size\PYZus{}t} \PY{n}{count}\PY{p}{;}
  \PY{p}{\PYZcb{}} \PY{n}{files}\PY{p}{;}
\PY{p}{\PYZcb{}}\PY{p}{;}
\PY{k}{struct} \PY{n}{dir\PYZus{}fileheader}\PY{p}{\PYZob{}}
  \PY{k}{struct} \PY{p}{\PYZob{}}
    \PY{k+kt}{uint8\PYZus{}t}\PY{o}{*}\PY{n}{elem}\PY{p}{;}
    \PY{k+kt}{size\PYZus{}t} \PY{n}{count}\PY{p}{;}
  \PY{p}{\PYZcb{}} \PY{n}{filename}\PY{p}{;}
  \PY{n}{fileentry} \PY{n}{contents}\PY{p}{;}
\PY{p}{\PYZcb{}}\PY{p}{;}\PY{c+cm}{/*..*/}
\PY{c+c1}{//The programmer calls these generated functions.}
\PY{k+kt}{int} \PY{n+nf}{gen\PYZus{}zip}\PY{p}{(}\PY{n}{NailArena} \PY{o}{*}\PY{n}{tmp\PYZus{}arena}\PY{p}{,}\PY{n}{NailStream} \PY{o}{*}\PY{n}{out}\PY{p}{,}\PY{n}{zip} \PY{o}{*}\PY{n}{val}\PY{p}{)}\PY{p}{;}
\PY{n}{zip}\PY{o}{*}\PY{n+nf}{parse\PYZus{}zip}\PY{p}{(}\PY{n}{NailArena} \PY{o}{*}\PY{n}{arena}\PY{p}{,} 
              \PY{k}{const} \PY{k+kt}{uint8\PYZus{}t} \PY{o}{*}\PY{n}{data}\PY{p}{,} \PY{k+kt}{size\PYZus{}t} \PY{n}{size}\PY{p}{)}\PY{p}{;}
\PY{c+c1}{//The programmers implements these two transformation}
\PY{c+c1}{//functions and two similar ones for output.}
\PY{k}{extern} \PY{k+kt}{int} \PY{n+nf}{zip\PYZus{}eod\PYZus{}parse}\PY{p}{(}\PY{n}{NailArena} \PY{o}{*}\PY{n}{tmp}\PY{p}{,}
 \PY{n}{NailStream} \PY{o}{*}\PY{n}{filestream}\PY{p}{,} \PY{n}{NailStream} \PY{o}{*}\PY{p}{,}\PY{n}{NailStream} \PY{o}{*}\PY{n}{current}\PY{p}{)}\PY{p}{;}
\PY{k}{extern} \PY{k+kt}{int} \PY{n+nf}{zip\PYZus{}compression\PYZus{}parse}\PY{p}{(}\PY{n}{NailArena} \PY{o}{*}\PY{n}{tmp}\PY{p}{,}
 \PY{n}{NailStream} \PY{o}{*}\PY{n}{uncomp}\PY{p}{,}\PY{n}{NailStream} \PY{o}{*}\PY{n}{compressed}\PY{p}{,}\PY{k+kt}{uint32\PYZus{}t}\PY{o}{*} \PY{n}{size}\PY{p}{)}\PY{p}{;}
\end{Verbatim}

>>>>

\begin{Verbatim}[commandchars=\\\{\},codes={\catcode`\$=3\catcode`\^=7\catcode`\_=8}]
\PY{k}{struct} \PY{n}{zip}\PY{p}{\PYZob{}}
        \PY{n}{end\PYZus{}of\PYZus{}directory} \PY{n}{contents}\PY{p}{;}
\PY{p}{\PYZcb{}}\PY{p}{;}
\PY{k}{struct} \PY{n}{end\PYZus{}of\PYZus{}directory}\PY{p}{\PYZob{}}
\PY{k+kt}{uint16\PYZus{}t} \PY{n}{disks}\PY{p}{;}
\PY{k+kt}{uint16\PYZus{}t} \PY{n}{directory\PYZus{}disk}\PY{p}{;}\PY{c+cm}{/*...*/}
\PY{k}{struct} \PY{p}{\PYZob{}}
        \PY{n}{dir\PYZus{}fileheader}\PY{o}{*} \PY{n}{elem}\PY{p}{;}
        \PY{k+kt}{size\PYZus{}t} \PY{n}{count}\PY{p}{;}
\PY{p}{\PYZcb{}} \PY{n}{files}\PY{p}{;}
\PY{p}{\PYZcb{}}\PY{p}{;}
\PY{k}{struct} \PY{n}{dir\PYZus{}fileheader}\PY{p}{\PYZob{}}
        \PY{k}{struct} \PY{p}{\PYZob{}}
                \PY{k+kt}{uint8\PYZus{}t}\PY{o}{*}\PY{n}{elem}\PY{p}{;}
                \PY{k+kt}{size\PYZus{}t} \PY{n}{count}\PY{p}{;}
        \PY{p}{\PYZcb{}} \PY{n}{filename}\PY{p}{;}
        \PY{n}{fileentry} \PY{n}{contents}\PY{p}{;}
\PY{p}{\PYZcb{}}\PY{p}{;}\PY{c+cm}{/*..*/}
\PY{c+c1}{//The programmer calls these functions to interact with the Nail\PYZhy{}generated parser.}
\PY{k+kt}{int} \PY{n+nf}{gen\PYZus{}zip}\PY{p}{(}\PY{n}{NailArena} \PY{o}{*}\PY{n}{tmp\PYZus{}arena}\PY{p}{,}\PY{n}{NailStream} \PY{o}{*}\PY{n}{out}\PY{p}{,}\PY{n}{zip} \PY{o}{*} \PY{n}{val}\PY{p}{)}\PY{p}{;}
\PY{n}{zip}\PY{o}{*}\PY{n+nf}{parse\PYZus{}zip}\PY{p}{(}\PY{n}{NailArena} \PY{o}{*}\PY{n}{arena}\PY{p}{,} \PY{k}{const} \PY{k+kt}{uint8\PYZus{}t} \PY{o}{*}\PY{n}{data}\PY{p}{,} \PY{k+kt}{size\PYZus{}t} \PY{n}{size}\PY{p}{)}\PY{p}{;}
\PY{c+c1}{//The programmer implements these two functions, and Nail calls them.}
\PY{k}{extern} \PY{k+kt}{int} \PY{n+nf}{zip\PYZus{}eod\PYZus{}parse}\PY{p}{(}\PY{n}{NailArena} \PY{o}{*}\PY{n}{tmp}\PY{p}{,}\PY{n}{NailStream} \PY{o}{*}\PY{n}{str\PYZus{}filestream}\PY{p}{,}
   \PY{n}{NailStream} \PY{o}{*}\PY{n}{str\PYZus{}end\PYZus{}directory}\PY{p}{,}\PY{n}{NailStream} \PY{o}{*}\PY{n}{current}\PY{p}{)}\PY{p}{;}
\PY{k}{extern} \PY{k+kt}{int} \PY{n+nf}{zip\PYZus{}compression\PYZus{}parse}\PY{p}{(}\PY{n}{NailArena} \PY{o}{*}\PY{n}{tmp}\PY{p}{,}\PY{n}{NailStream} \PY{o}{*}\PY{n}{str\PYZus{}uncompressed}\PY{p}{,}
   \PY{n}{NailStream} \PY{o}{*}\PY{n}{compressed}\PY{p}{,}\PY{k+kt}{uint32\PYZus{}t}\PY{o}{*} \PY{n}{size}\PY{p}{)}\PY{p}{;}
\end{Verbatim}

### 2015-08-10 paper/thesis/eval.tex
thesis

In Section~\ref{s:design}, we introduced Nail's syntax with a grammar
for DNS packets, shown in Figure~\ref{fig:dns-xform}.
The grammar corresponds almost directly to the diagrams in RFC 1035,
which defines DNS~\cite[\S4]{RFC:1035}.  Each DNS packet consists of a
header, a set of question records, and a set of answer records. Domain
names in both queries and answers are encoded as a sequence of labels,
terminated by a zero byte.  Labels are Pascal-style strings, consisting
of a length field followed by that many bytes comprising the label.

@@

Handling label compression in existing tools, such as Bison or Hammer,
would at best be very awkward, because some ad-hoc trick would have to
be used to re-position the parser's input stream.  Keeping track of the
position of all recognized labels would not be enough, as the offset field
may refer to any byte within the packet, not just the beginning of labels.
For this reason, the DNS server used as the example for Hammer does not
support compression.

@@

The \cc{zip\_file} grammar first splits the entire file stream into two
streams based on the \cc{zip\_end\_of\_directory} transform on line 2.
%(whose two C function signatures are shown in Figure~\ref{fig:zip-eod-xform}). 
\cc{zip\_end\_of\_directory_parse}
finds the end-of-directory header as described above, by scanning the
file backwards, and splits the file into two streams, one containing
the end-of-directory header and one containing the file contents. The
\cc{end\_of\_directory} rule is then applied to the header stream in
line 4. All offsets in the ZIP file refer to the beginning of the
file, so the stream \cc{\$file} which contains the file contents
without the header is passed as an argument to all parsers from
hereon.

@@

To evaluate how much programmer effort is required to build an
application that uses Nail, we implemented two applications---a
DNS server, and an
\cc{unzip} program---based on the above grammars, and compared code
size with comparable applications that process data manually, using
\cc{sloccount}~\cite{sloccount}.  We also compare the code size of our
DNS server to a DNS server written using the Hammer parsing framework,
although it does not fully support DNS (e.g., it lacks label compression,
among other things).  Figure~\ref{fig:effort} summarizes the results.

\begin{figure}[tb]
\centering
\smaller[0.5]
\begin{tabular}{@{~}lrr@{~}l@{~}}
\toprule
\textbf{Application}
  & \textbf{LoC w/ Nail}
  & \multicolumn{2}{c}{\textbf{LoC w/o Nail}} \\
\midrule
DNS server
  & 295
  & 683
  & (Hammer parser) \\

@@

In \S\ref{s:motivation}, we found 15 input handling vulnerabilities
related to ZIP files.

11 of these vulnerabilities involved memory corruption during input handling.
Because Nail's generated code checks offsets before reading and does not
expose any untrusted pointers to the application, it is immune against
memory corruption attacks by design.

Nail also protects against parsing inconsistency vulnerabilities like the four others we studied. First,
Nail grammars explicitly encode duplicated information such as the redundant length fields in ZIP
that caused a vulnerability in the Python ZIP library. Second, the other three vulnerabilities exist
because multiple implementations of the same protocol disagree on some inputs. Hand-written protocol
parsers are not very reusable, as they build application specific data structures and are tightly
coupled to the rest of the code. Nail grammars, however, can be re-used between applications,
avoiding protocol misunderstandings.

@@

The results are shown in Figure~\ref{fig:perf-dns}, and demonstrate that
our Nail-based DNS server can achieve higher performance and lower latency
than BIND\@.  Although BIND is a more sophisticated DNS server,
and implements many features that are not present in our Nail-based DNS
server and allow it to be used in more complicated configurations, 
we believe our results demonstrate that Nail's parsers are not
a barrier to achieving good performance.

>>>>

In Section~\ref{s:design}, we introduced Nail's syntax with a grammar
for DNS packets, shown in Figure~\ref{fig:dns-full}.
The grammar corresponds almost directly to the diagrams in RFC 1035,
which defines DNS~\cite[\S4]{RFC:1035}.  Each DNS packet consists of a
header, a set of question records, and a set of answer records. Domain
names in both queries and answers are encoded as a sequence of labels,
terminated by a zero byte.  Labels are Pascal-style strings, consisting
of a length field followed by that many bytes comprising the label.

@@

Handling label compression in existing tools, such as Bison or Hammer,
would be awkward at best, because some ad-hoc trick would have to
be used to re-position the parser's input stream.  Keeping track of the
position of all recognized labels would not be enough, as the offset field
may refer to any byte within the packet, not just the beginning of labels.
For this reason, the DNS server used as the example for Hammer does not
support compression.

@@

The \cc{zip\_file} grammar first splits the entire file stream into two
streams based on the \cc{zip\_end\_of\_directory} transform on line 2.
%(whose two C function signatures are shown in Figure~\ref{fig:zip-eod-xform}). 
The corresponding C function
\cc{zip\_end\_of\_directory_parse}
finds the end-of-directory header as described above, by scanning the
file backwards, and splits the file into two streams, one containing
the end-of-directory header and one containing the file contents. The
\cc{end\_of\_directory} rule is then applied to the header stream in
line 4. All offsets in the ZIP file refer to the beginning of the
file, so the stream \cc{\$file} which contains the file contents
without the header is passed as an argument to all parsers from
hereon.

@@

To evaluate how much programmer effort is required to build an
application that uses Nail, we implemented two applications---a
DNS server and an
\cc{unzip} program---based on the above grammars, and compared code
size with comparable applications that process data manually, using
\cc{sloccount}~\cite{sloccount}.  We also compare the code size of our
DNS server to a DNS server written using the Hammer parsing framework,
although it does not fully support DNS (e.g., it lacks label compression,
among other things).  Figure~\ref{fig:effort} summarizes the results.

\begin{figure}[tb]
\centering
\smaller[0.5]
\begin{tabular}{lrr@{~}l}
\toprule
\textbf{Application}
  & \textbf{LoC w/ Nail}
  & \multicolumn{2}{c}{\textbf{LoC w/o Nail}} \\
\midrule
DNS server
  & 295
  & 683
  & (Hammer parser) \\

@@

In \S\ref{s:motivation}, we presented 15 input handling vulnerabilities
related to ZIP files.

11 of these vulnerabilities involved memory corruption during input handling.
Because Nail's generated code checks offsets before reading and does not
expose any untrusted pointers to the application, it is immune to
memory corruption attacks by design.

Nail also protects against parsing inconsistency vulnerabilities like the four others we studied.
Nail grammars explicitly encode duplicated information such as the redundant length fields in ZIP
that caused a vulnerability in the Python ZIP library. The other three vulnerabilities exist
because multiple implementations of the same protocol disagree on some inputs. Hand-written protocol
parsers are not very reusable, as they build application-specific data structures and are tightly
coupled to the rest of the code. Nail grammars, however, can be re-used between applications,
avoiding protocol misunderstandings.

@@

The results are shown in Figure~\ref{fig:perf-dns}, and demonstrate that
our Nail-based DNS server can achieve higher performance and lower latency
than BIND\@.  Although BIND is a more sophisticated DNS server,
and implements many features that are not present in our Nail-based DNS
server and that allow it to be used in more complicated configurations, 
we believe our results demonstrate that Nail's parsers are not
a barrier to achieving good performance.

### 2015-08-10 paper/thesis/impl.tex
thesis

The current prototype of the Nail parser generator supports the C programming
language. The implementation parses Nail grammars with Nail itself, using a
130-line Nail grammar feeding into a 2,000-line C++ program that
emits the parser and generator code. Bootstrapping is
performed with a subset of the grammar implemented using conventional grammars. 
An option for C++ STL data models is in development.  In
this section, we will discuss some particular features of our parser
implementation.

A generated Nail parser makes two passes through the input: the first to
validate and recognize the input, and the second to bind this data to the internal
model. Currently the parser uses a straightforward top-down algorithm, which can perform poorly on
grammars that backtrack heavily. However, preparations have been made to add Packrat
parsing~\cite{packrat-parsing:icfp02} that achieve linear time even in the worst case.

>>>>

The current prototype of the Nail parser generator can emit parsers in either C or C++. The implementation parses Nail grammars with Nail itself, using a
130-line Nail grammar feeding into a 2,000-line C++ program that
emits the parser and generator code. Bootstrapping is
performed with a subset of the grammar implemented using conventional grammars. 
In this section, we will discuss some particular features of our parser
implementation.

A generated Nail parser makes two passes through the input: the first to
validate and recognize the input, and the second to bind this data to the internal
model. Currently the parser uses a straightforward top-down algorithm, which can perform poorly on
grammars that backtrack heavily. However, Packrat
parsing~\cite{packrat-parsing:icfp02} that achieves linear time even in the worst case can be added relatively easily.

\paragraph{Parser input/output.}
Nail-generated parsers interact with their input through an abstract stream data type, which must support the operations show in Figure~\ref{fig:stream-api}. In the C implementation, the programmer can provide an implementation of these operations, or fall back to using a default implementation backed by continuous buffers in memory. Both the input stream and all intermediate streams must use the same implementation.

While the concrete implementation is transparent to the Nail parser itself, transformations can and often must modify the stream data type through other interfaces. The standard library of Nail transformations is at present only for memory-buffer backed streams and often directly manipulates the underlying buffers. This makes many operations, such as splitting a stream into multiple substreams, much faster than if the substreams had to be copied directly.

\input{fig-stream-api}
\paragraph{C++ implementation.}
The C implementation forces all streams to have the same underlying data type. To address this shortcoming, we developed an optional C++ backend for Nail. Nail C++ parsers are template functions that are parametrized by the types of the streams they operate on. Transformations are template functors -- classes that can perform different computation  and produce a different type of stream depending on the types of their inputs.

Useful transformations are often only possible to write when specialized to their input types, and the fully general will either be very slow or not even implemented. Consider for example the standard library offset transform, which creates a substream starting at a given offset. Given just the stream interface shown in Figure~\ref{fig:stream-api}, the only way to implement offset would be to either to make a copy of the underlying stream, using reposition and read, or to reposition the underlying stream before and after each read. By specializing the transformation for in-memory streams, we can construct another memory stream that points to the correct range in the same memory. Similar specializations could exist for file-descriptor backed streams using a duplicate descriptor or sparse-fragment streams.

Transformations can also be expressed through chaining together other transformations -- which is equivalent to composing them directly in the Nail grammar, but results in shorter and easier to understand Nail grammars.

In addition to allowing complicated file formats to be captured more naturally by mapping their layout onto different stream types, we will show in Section~\ref{sec:network-stack} that templated streams allow new classes of applications for Nail transformations, such as reassembling packets.

### 2015-08-10 paper/thesis/relwk.tex
thesis

Closely related work has been done in the field of data description languages, for example
PacketTypes~\cite{mccann2000packet} and DataScript~\cite{back2002datascript}. 
A broader overview of data
description languages can be found in~\cite{Fisher:700DDL}. However, PacketTypes works
only as a parser, not as an output generator and implement a C-like structure model enhanced with
length fields and constraints. DataScript adds output generation and built-in support for offset
fields. A
particularly sophisticated data description language, PADS~\cite{Fisher:2005:PDL:1064978.1065046},
which is targeted more towards offline analysis, even features built-in support for XML and
automatic grammar inference.

However, these systems cannot easily handle complicated encodings such as compressed data, which are
supported by Nail's stream transforms. While sophisticated languages like PADS allow for handling
particular variations of offset fields, compressed data or even XML entities, each of these
features has to be implemented in the data description language - and all associated tools. Nail's
transformations keep the core language practicably small, while enabling the wide range of
features real-world protocols require.

@@

Parsifal~\cite{ANSSI:parsifal} is a parser framework that also supports
generating output for OCaml.  Parsifal structures grammars as an OCaml
type that holds an internal model and functions for parsing input and
output.  However, Parsifal can produce parsers and generators
only for simple, fixed-size structures.  The programmer can then use
these when implementing parsers and generators for more complicated
formats, manually handling offsets, checksums, and the like, risking bugs.
Nail handles more complicated constructs without the programmer manually
writing code to support them.

>>>>

Closely related work has been done in the field of data description languages, for example
PacketTypes~\cite{mccann2000packet} and DataScript~\cite{back2002datascript}; a
broader overview of data description languages can be found in Fisher et
al~\cite{Fisher:700DDL}.
PacketTypes implements a C-like structure model enhanced with
length fields and constraints, but works only as a parser, and
not as an output generator.
DataScript adds output generation and built-in support for offset fields.
A particularly sophisticated data description language, PADS~\cite{Fisher:2005:PDL:1064978.1065046},
which is targeted more towards offline analysis, even features built-in support for XML and
automatic grammar inference.
However, these systems cannot easily handle complicated encodings such as compressed data, which are
supported by Nail's stream transforms. While sophisticated languages like PADS allow for handling
particular variations of offset fields, compressed data, or even XML entities, each of these
features has to be implemented in the data description language and all associated tools. Nail's
transformations keep the core language small, while enabling the wide range of
features real-world protocols require.

@@

Parsifal~\cite{ANSSI:parsifal} is a parser framework for OCaml that also supports
generating output.  Parsifal structures grammars as an OCaml
type that holds an internal model and functions for parsing input and
output.  However, Parsifal can produce parsers and generators
only for simple, fixed-size structures.  The programmer can then use
these when implementing parsers and generators for more complicated
formats, manually handling offsets, checksums, and the like, risking bugs.
Nail handles more complicated constructs without the programmer manually
writing code to support them.
